---
output:
  word_document: 
    keep_md: yes
  html_document:
    keep_md: yes
---
Reproducible Research Week 2 Assignment
=======================================

Loading necessary packages and downloading and reading in the necessary data:
```{r, results='markup'}
library(knitr)
library(tidyr)
library(dplyr)
path <- "t:/pd/rscripts/activity.csv"
#download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", path)
activity <- read.csv(path)
```



Question: What is the mean total number of steps taken per day?

First we will remove the NAs from the data set, then group by date and summarise to calculate the total / mean / median steps per day:
```{r, results='markup'}
activityNoNA <- activity[!is.na(activity$steps),]

activityNoNAsteps <- activityNoNA%>%
    group_by(date)%>%
    summarise(totalsteps = sum(steps), meansteps = mean(steps), mediansteps = median(steps))

activityNoNAsteps[,c("date", "totalsteps")]
```

Creating a histogram of the total number of steps taken per day:
```{r, results='markup'}
hist(activityNoNAsteps$totalsteps, xlab = "Total Steps Per Day")
```

Calculating and reporting the mean and median of the total number of stapes taken per day:
```{r, results='markup'}
Meanstepsperday <- mean(activityNoNAsteps$totalsteps)
Medianstepsperday <- median(activityNoNAsteps$totalsteps)
Meanstepsperday
Medianstepsperday
```



Question: What is the average daily pattern?

First I will group by the interval and summarize by mean steps per day, then create a time series plot of the average steps per interval across  days:
```{r, results='markup'}
activityNoNAintervals <- activityNoNA%>%
    group_by(interval)%>%
    summarise(meanintervalsteps = mean(steps))

with(activityNoNAintervals, plot(interval, meanintervalsteps, type = "l", xlab = "Interval", ylab = "Avg of Mean Interval Steps Across Days"))
```

Calculating and reporting the interval which contains the max number of steps:
```{r, results='markup'}
Maxinterval <- which.max(activityNoNAintervals$meanintervalsteps)
activityNoNAintervals[Maxinterval, ]
```



Question: Imputing missing values 

Subsetting by only rows with missing data then calculating and reporting the number of missing values:
```{r, results='markup'}
activityNAsteps <- activity[is.na(activity$steps),]
nrow(activityNAsteps)
```

I replace the missing values with the average steps for the given interval:
```{r, results='markup'}
activityimputed <- left_join(activity, activityNoNAintervals, by='interval') %>%
    mutate(steps = ifelse(is.na(steps), meanintervalsteps, steps))
```

Creating a histogram of the total number of steps taken each day for the data frame including imputed data:
```{r, results='markup'}
activityimputed_date <- activityimputed%>%
  group_by(date)%>%
  summarise(totalsteps = sum(steps), meansteps = mean(steps), mediansteps = median(steps))

hist(activityimputed_date$totalsteps)
abline(v = mean(activityimputed_date$totalsteps), col = "red", lwd = 2)
```

Calculating the mean and median of total average steps using imputed data where NA: 
```{r}
mean(activityimputed_date$totalsteps)
median(activityimputed_date$totalsteps)
```

Then compare them to the figures computed without imputed values to see if there is any difference:
```{r, results='markup'}
identical(mean(activityNoNAsteps$totalsteps), mean(activityimputed_date$totalsteps))
identical(median(activityNoNAsteps$totalsteps), median(activityimputed_date$totalsteps))
```

Mean does not change when you impute values however the median does slighty:
```{r, results='markup'}
median(activityimputed_date$totalsteps) - median(activityNoNAsteps$totalsteps) 
```
The median for total steps is 1.18 greater for the imputed data than that with the NAs removed, a meaningless figure.



Question: Are there differences in activity patterns between weekdays and weekends?

Creating a new factor for weekends vs weekdays:
```{r, results='markup'}
activityimputed$date <- as.Date(activityimputed$date)

activityimputed <- activityimputed %>% 
  mutate(weekday = weekdays(date))

weekday_weekend = c("Monday" = "Weekday", "Tuesday" = "Weekday", "Wednesday" = "Weekday", "Thursday" = "Weekday", "Friday"="Weekday", "Saturday" = "Weekend", "Sunday"="Weekend") 
activityimputed$weekend <- as.factor(weekday_weekend[activityimputed$weekday])

activityimputed_byintervalweekend <- activityimputed %>%   
  filter(weekend == "Weekend") %>% 
  group_by(interval) %>% 
  summarise(totalsteps = sum(steps), meansteps = mean(steps), mediansteps = median(steps))

activityimputed_byintervalweekday <- activityimputed %>%   
    filter(weekend == "Weekday") %>% 
    group_by(interval) %>% 
    summarise(totalsteps = sum(steps), meansteps = mean(steps), mediansteps = median(steps))
```


Plotting the difference in activity for weekend and weekdays:
```{r, results='markup'}

par(mfrow = c(2, 1), mar = c(4,4,4,4))
rng <- range(activityimputed_byintervalweekday$meansteps)
with(activityimputed_byintervalweekday, plot(interval, meansteps, type = "l", col = "blue", xlab = "Mean Steps Per Interval- Weekdays", ylim = rng))
with(activityimputed_byintervalweekend, plot(interval, meansteps, type = "l", col = "red", xlab = "Mean Steps Per Interval - Weekends", ylim = rng))
```

and together on a single plot:
```{r, results='markup'}
par(mfrow = c(1, 1))
with(activityimputed_byintervalweekday, plot(interval, meansteps, type = "l", col = "blue", xlab = "Mean Steps Per Interval", ylim = rng))
with(activityimputed_byintervalweekend, points(interval, meansteps, type = "l", col = "red", ylim = rng))
legend("topright", pch = 19, col = c("blue", "red"), legend = c("Weekdays", "Weekends"))
```

